---
title: "Injustice in Our Justice System"
author: "Ethan Side"
date: '2022-11-01'
output: html_document
---

```{r echo=FALSE, eval=TRUE, message=FALSE}
source("../source/analysis.R")
```

## Abstract 
The US justice system has been something that for centuries has set the country 
apart yet over the last couple decades it has filled the prison and jail systems
in an inequitable way. The following visualizations can shed some light n these 
injustices. In hopes that awareness can be spread to help promote equity for all.




## Introduction 
The Vera Institute allows viewers of their data set to delve
into decades worth of jail and prison populations from the United States
of America and the states and counties within. The Vera Institute also
allows us to see the total populations certain genders and races so that
comparisons can be made in order for us to see possible injustices in
the United States justice system. Direct stakeholders for a data set
like this are policy makers, socially aware citizens. They can access a
data set like this and cause change in society. A policy maker can see a
march by a citizen and make change based off this kind of data.
Indirectly those who are incarcerated might be able to be set free if
certain laws were passed, and also indirectly would be every US citizen
as their justice system could start to become more equitable if the
direct stakeholders are able to harness the data in the right way. Some
of the questions I will answer in this report are: What are certain
states jail populations over a quarter decade? How has the US Prison
population fluctuated over a quarter decade? What counties in Alabama
disenfranchise African-Americans?

## Data Summary 
Summary In this report we cover total US prison populations
all the way down to certain county statistics to get to know the US
prison system on a large and small scale. Despite not being the second
most populous race in Washington they are the second most jailed, the
*black* population in Washington is second in all counties except 4. The
most inequitable county in Alabama is *Winston* with an 11.08 inequality
rate. 11 times more inequitable than an equal rate of 1.00. Which year
did the prison populations of Washington, Mississippi, Illinois, and
Pennsylvania reach its peak? *1999*, this could be a culmination of
things but race riots over the last decade in a massive city like
Chicago would be a reason for this year being the peak.



## Growth of U.S. Prison Population
```{r}

chart3

```

__United States prison population over a 26 year period from 1990 to 2016__

The key takeaway from this graph is the steady upward trend until the
change of the millennium with a small decrease in 2000 and 2001. The
graph continues almost a plateau for the next decade until 2012 when a
steep decrease in the prison population starts to occur. Why was there a
steady decrease after 2012? This could be due to the legalization of
marijuana around this time in the states of Washington and Colorado.
Decriminalization of the drug became popularized at this time as well.


# Section 4: Growth of Prison Population by State

```{r}

chart4
```

__Prison Populations from 1990 to 2016 in Washington, Mississippi,__
__Illinois, and Pennsylvania__

Why the stark difference in incarcerated
totals? These four states all in different regions of the United States
have differing state populations and answer the question of the large
range between each state. Another sore thumb in this data set is the
massive upswing in incarcerations in Illinois during the 90s. This could
be because of the general growth of the state's population but is more
likely due to race riots happening during that era and more brutal
police enforcement, especially in large urban areas like Chicago. This
uptick is noticeable in Pennsylvania too.


# Section 5: Washington State Jail Distribution is Unequitable

```{r}
plot5
```

__Washington counties max jail population by race in 2018__


When it comes to each counties max jail population in the latest year we
have data for what is the order that the races come in? In other words
is there a consistent pattern of races when it comes to the hierarchy of
jail populations. As it turns out in the state of Washington most
counties, in descending order, tend to go from white, black, latinx, to
Asian-american pacific islanders. This ordering is intriguing as latinx
populations in Washington are second only to the white population. In
counties like Yakima and Skagit this swap between the black max and the
latinx max exist resulting in what should be a better representation of
the Washington population. The black community being the second highest
prison populator is a major inequality in the state.

# Section 6: African-American Inequality in the Jail Population of Alabama Counties

```{r}
plot6
```

__Alabama racial inequalities toward African-Americans over the last 30 years by county__

Which counties in a deep south state like Alabama don't have racial
inequalities in their jail toward African Americans? To answer this
question we refer to the legend on the map with a lighter color meaning
harsher inequalities and darker meaning the jail population of
African-Americans more closely represents the counties population of
African-Americans. Sadly the answer to the previous question is there
are no counties in Alabama that don't have some kind of injustice when
it comes to the jailing of African-Americans. While some are close to
the magic number of 1 every county is above that equality standard
making every county having jailing biases.

